---
title: "Assignment 3: Data Exploration"
author: "Gary Alvarez"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
getwd() #I checked that my Working directory is set to the base folder for 
#the Environmental Data Analytics Course repository.
#install.packages("lubridate"), tidyverse and here.
library(lubridate)
library(tidyverse)
library(here)
library(ggplot2)

Neonics <- read.csv(
  "./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",
  stringsAsFactors = TRUE
  )   
Litter <- read.csv(
  "./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",
  stringsAsFactors = TRUE
  )   
#I uploaded the two datasets.
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Studying the ecotoxicology of neonicotinoids on insects is essential for balancing the benefits of pest control in agriculture with the potential risks to biodiversity, ecosystem functioning, and human well-being. It helps guide sustainable agricultural practices and the development of alternative pest management strategies.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Studying litter and woody debris in forest ecosystems provides a comprehensive understanding of the ecological processes that shape these environments. This knowledge is critical for sustainable forest management, conservation efforts, and predicting the impacts of environmental changes on ecosystems.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and fine woody debris are collected from elevated and ground traps, respectively.
 2. Ground traps are sampled once per year. Target sampling frequency for elevated traps varies by vegetation present at the site, with frequent sampling (1x every 2weeks) in deciduous forest sites during senescence, and in-frequent year-round sampling (1x every 1-2 months) at evergreen sites. 
 3. In this protocol, litter is defined as material that is dropped from the forest canopy and has a butt end diameter <2cm and a length <50 cm; this material is collected in elevated 0.5m2 PVC traps. Fine wood debris is defined as
material that is dropped from the forest canopy and has a butt end diameter <2cm and a length >50 cm; this material is collected in ground traps as longer material is not reliably collected by the elevated traps.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Summary of Neonics}
class(Neonics) #The type of file is data frame.
colnames(Neonics) #We can see the names of the 30 columns that make up the data frame.
dim(Neonics) #Neonics has 4623 rows and 30 columns.
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Most Common Effects of Neonics}
summary(Neonics$Effect)
```

> Answer:The three most common effects studied are Population, Mortality, and Behavior. These effects might specifically be of interest because they offer information about the population size of insects at a given time (Population), the rate at which they are dying (Mortality), and the general behavior they express (Behavior).

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Most Commonly Studied Species}
sum_common_names <- summary(Neonics$Species.Common.Name)
sort(sum_common_names, decreasing = TRUE) #Sorting the output of the summary line

```

> Answer: Excluding the category Other (670), the six most commonly studied species in the dataset are: 1. Honey Bee (667), 2. Parasitic Wasp (285), 3. Buff Tailed Bumblebee (183), 4. Carniolan Honey Bee (152), 5. Bumble Bee (140), 6. Italian Honeybee (113). All these species are pollinators, and they are more important over other species because the insecticides kill all insects within a range of land. That includes pollinators, whose work is benefitial to the ecosystems.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Concentrations}
class(Neonics$Conc.1..Author)
summary(Neonics$Conc.1..Author)
```

> Answer: The class of `Conc.1..Author.` is factor, because it represents categories.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Visualization of Studies Conducted by Publication Year}

ggplot(Neonics)  +
geom_freqpoly(aes(x = Publication.Year))

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Studies Conducted by Publication Year by Test Locations}
ggplot(Neonics)  +
geom_freqpoly(aes(x = Publication.Year, color = Test.Location))

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are Lab and Fielt natural. They change over time. From 1990 to 2000, Field artificial and Lab were the most common locations. However, after a peak of Field natural in 2009, this location became significantly less common than Lab, and this trends holds until 2020.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Endpoint counts}
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer: The most common endpoints are NOEL and LOEL. The terms "NOEL" (No Observed Effect Level) and "LOEL" (Lowest Observed Effect Level) are used in toxicology and risk assessment to describe exposure levels at which no adverse effects are observed (NOEL) or the lowest exposure level at which adverse effects are observed (LOEL).

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Litter Samples in 2018}
class(Litter$collectDate) #The class of collectDate is originally "factor".
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
unique(Litter$collectDate, incomparables = FALSE)
```

> Answer: The dates litter was sampled in August 2018 are August 02 and August 30.

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Plots Sampled at Niwot Ridge}
unique(Litter$plotID, incomparables = FALSE)
summary(Litter$plotID)
```

> Answer: 12 plots were sampled at Niwot Ridge. The 'unique' function is used to obtain the unique values of a vector or a column in a data frame. It returns a vector containing the unique elements in the order in which they appear in the input. On the other hand, the 'summary' function provides a summary of the distribution of a variable, typically for numerical variables. Since plotID is not a numeric variable, it is better to use the 'unique' function.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Bar graph of Functional Groups}
ggplot(Litter, aes(x =functionalGroup))+
       geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Box plot vs. Violin plot of Dry Mass by Functional Group}
ggplot(Litter)+
       geom_boxplot(aes(x =functionalGroup, y=dryMass))

ggplot(Litter)+
       geom_violin(aes(x =functionalGroup, y=dryMass))
```

16. Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: Because the boxplot is able to convey more meaningful information about the distribution. With the characteristics of this dataset, the violin plot looks like multiple vertical lines, and that doesn't provide much information.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles, and Twigs/branches.
