---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Gary Alvarez Mejia"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}

#1
library(tidyverse)
library(agricolae)   
library(here)
#install.packages("corrplot")
library(corrplot)
library(dplyr)
here()

Lake_ChemPhys <- read.csv(here(
  "Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
  stringsAsFactors = TRUE)
Lake_ChemPhys$sampledate <- as.Date(
  Lake_ChemPhys$sampledate, format = "%m/%d/%y")

class(Lake_ChemPhys$sampledate)
#2
Garytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(Garytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer: temperature_C = B0 + B1*depth + e
H0: There is no significant relationship between the mean lake temperature recorded during July and the depth across all lakes. (B1 = 0)
Ha: There is a significant relationship between the mean lake temperature recorded during July and the depth across all lakes. (B1 != 0)


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
library(dplyr)
Lake_filtered <- Lake_ChemPhys %>%
  # Filter only dates in July
  filter(format(Lake_ChemPhys$sampledate, "%m") == "07") %>%
  # Select only the specified columns
  select(lakename, year4, daynum, depth, temperature_C) %>%
  # Remove rows with missing values (NAs)
  drop_na()


#5
Temp_by_Depth_plot <- ggplot(
  Lake_filtered, aes(x = depth, y = temperature_C)) +
  # Add points for the scatter plot
  geom_point() +
  # Add a smoothed line showing the linear model
  geom_smooth(
    method = "lm", se = FALSE, 
    color = "blue") +
  # Set limits for the temperature values from 0 to 35 °C
  ylim(0, 35) +
  # Add labels and title
  labs(x = "Depth",
       y = "Temperature (°C)",
       title = "Scatter Plot of Temperature by Depth") 
# Print the plot
print(Temp_by_Depth_plot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest anything about the linearity of this trend?

> Answer: The figure suggests that at surface level (depth=0), the temperatures are high, and as we go deeper (depth>0), the temperature decreases.
The distribution of points suggests that there is a weak linear trend that connects depth and temperature.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
depth_temp.regression <- lm(
  data = Lake_filtered,
  temperature_C ~ depth)
summary(depth_temp.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The results of the regression show that there is a negative relation between Temperature in Celsius and depth in meters. This means that, as meters increase (+), temperature decreases (-). According to the R-squared value, 73% of the variability in temperature is explained by changes in depth. The finding is based on 9726 degrees of freedom. The result is statistically significant (p-value < 0.05).
The result of the regression suggests that for every additional meter (+1m) of change in depth, the temperature decreases by 1.94° Celsius (-1.94°C).


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
allvariables_reg <- lm(data = Lake_filtered, temperature_C ~ depth + 
              year4 + daynum)
summary(allvariables_reg)

step(allvariables_reg)
#Since the AIC value didn't increase with 
#the removal of variables, I decided to keep them 
#all in the model

#10
best_model <- lm(
  data = Lake_filtered, temperature_C ~ depth + 
              year4 + daynum)
summary(best_model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests that we add the variables year4 and daynum to depth. If we keep year4 and daynum alongside with depth in the model, the model explains 74% of the variance in temperature_C (R-squared = 0.74). Comparing the results with the single-predictor regression (only depth), which showed a prediction of 73.8%, the model that includes all the variables is slightly better.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
anova_test_temp <- aov(
  data = Lake_filtered, temperature_C ~ lakename)
summary(anova_test_temp)

linear_model <- lm(
  temperature_C ~ lakename, data = Lake_filtered)
summary(linear_model)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, the mean temperature is different among the analyzed lakes, and that difference is statistically significant with a p-value < 0.05  (<2e-16).
The lineal model shows that, accross all lakes, there is a difference in mean temperature, and that difference is statistically significant, with a confidence level of 95%, considering all the t-values and p-values.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
temp_by_depth_bylake <-
  ggplot(
    Lake_filtered, aes(
      x=depth, y=temperature_C, color=lakename)) +
      geom_point(alpha=0.5, size= 1) +
  geom_smooth(method = lm, se = FALSE)+
  ylim(0, 35)+
  ggtitle("Temperature by Depth across lakes") +
   labs(x="Depth", y="C°")
print(temp_by_depth_bylake)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(anova_test_temp)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: According to the Tukey HSD test, Paul Lake and Ward Lake have the same mean temperature as Peter Lake, statistically speaking.
No, no other lake has a mean temperature that is statistically distinct from all other lakes.

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we're interested in comparing the mean temperatures of Peter Lake and Paul Lake to see if they are statistically distinct, we can use a two-sample t-test.

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
Crampton_Ward_Lakes <- Lake_filtered %>%
  filter(
    lakename %in% c("Crampton Lake", "Ward Lake"))

Crampton_Ward_twosample_t.test <- t.test(
  Crampton_Ward_Lakes$temperature_C ~ Crampton_Ward_Lakes$lakename)
print(Crampton_Ward_twosample_t.test)
```

>Answer: With a p-value of 0.2649 (greater than 0.05), and a t-statistic of 1.11 (smaller than 1.96), there is no evidence to reject the null hypothesis that the mean temperature between lakes is statistically the same.
It doesn't match my answer for part 16.

