---
title: "Assignment 10: Data Scraping"
author: "Gary Alvarez Mejia"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(lubridate)
library(here); here()

#install.packages("rvest")
library(rvest)

# Set theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
#Fetch the web resources from the URL
webpage <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
webpage
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water_system_name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water_system_name

PWSID <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max_day_use <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max_day_use

months <- webpage %>% 
  html_nodes(".fancy-table:nth-child(31) tr+ tr th") %>% 
  html_text()
months

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
df_withdrawals <- data.frame("Ownership" = ownership,
                             "PWSID" = PWSID,
                             "Water System Name" = water_system_name,
                             "Month" = rep(c(months)),
                             "Year" = rep(2022, each = 12),
                             "Max Day Use" = as.numeric(max_day_use))
                             
df_withdrawals <- df_withdrawals %>% 
  mutate(Date = my(paste(Month,"-",Year)))

#5 
ggplot(df_withdrawals,aes(x=Date,y=Max.Day.Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2022 Maximum Water day usage data for",water_system_name),
       y="Max Day Use (MGD)",
       x="Date") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape.it <- function(the_year, the_PWSID){

the_scrape_url <- paste0(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?', 
  'pwsid=', the_PWSID, '&', 'year=', the_year)
webpage <- read_html(the_scrape_url)

water_system_name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()

PWSID <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()

ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()

max_day_use <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

df_withdrawals_func <- data.frame("Ownership" = ownership,
                             "PWSID" = PWSID,
                             "City" = water_system_name,
                             "Month" = months,
                             "Year" = rep(the_year, each = 12),
                             "Max Day Use" = as.numeric(max_day_use))

return(df_withdrawals_func)
}

#Trying out the function
scrape.it(2014, "01-11-010")

df_withdrawals_q6 <- scrape.it(2014, "01-11-010")

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
scrape.it(2015, "03-32-010")

df_withdrawals_q7 <- scrape.it(2015, "03-32-010")

df_withdrawals_q7 <- df_withdrawals_q7 %>% 
  mutate(Date = my(paste(Month,"-",Year)))

ggplot(df_withdrawals_q7,aes(x=Date,y=Max.Day.Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste(2015, "Maximum Water day usage data for", "Durham"),
       y="Max Day Use (MGD)",
       x="Date") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
scrape.it(2015, "01-11-010")

df_withdrawals_q8 <- scrape.it(2015, "01-11-010")

df_withdrawals_q8 <- df_withdrawals_q8 %>% 
  mutate(Date = my(paste(Month,"-",Year)))

ggplot(df_withdrawals_q8,aes(x=Date,y=Max.Day.Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste(2015, "Maximum Water day usage data for", "Asheville"),
       y="Max Day Use (MGD)",
       x="Date") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")


df_combined <- bind_rows(df_withdrawals_q7, df_withdrawals_q8)

df_combined

ggplot(df_combined,aes(x=Date,y=Max.Day.Use, color=City)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste(2015, "Maximum Water day usage data for Durham and Asheville"),
       y="Max Day Use (MGD)",
       x="Date") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
Asheville_decade <- 2010:2021
Asheville_id <- rep('01-11-010', length(Asheville_decade))

df_Asheville <- map2(Asheville_decade, Asheville_id, scrape.it)
df_Asheville <- bind_rows(df_Asheville)

df_Asheville <- df_Asheville %>% 
  mutate(Date = my(paste(Month,"-",Year)))

Asheville_plot <- ggplot(df_Asheville, aes(x=Date,y=Max.Day.Use)) +
  geom_line() +
  geom_smooth(method = 'loess', se = F, color = 'blue') +
  labs(x = "Date", y = "Maximum Daily Withdrawals") +
  theme_light() +
  ggtitle("Maximum Daily Withdrawals in Asheville by month: 2010-2021")
Asheville_plot

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: Yes, Asheville has a trend in water usage over time. The plot shows that from 2010 to 2015 the water usage slighltly decreased, but it started increasing significantly after 2015 (and in the 2015-2021 period). 
>
